The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-07 rational.tex

\begin{comment}
%%%%%%%%%%%%%%%%%%
\subsection{Syntax}

%subcode{bnf} include main
% Values, v = () | (v, v) | left v | right v
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
%
% Sequential Contexts, P, F = [] | P:c
% Parallel Contexts, D = [] 
%                     &|& D:(P| [](+)c| F)
%                     &|& D:(P| c(+)[]| F)
%                     &|& D:(P| [] (*) c v| F)
%                     &|& D:(P| [] (*) v c| F)
%                     &|& D:(P| c v (*)  []| F)
%                     &|& D:(P| v c (*) []| F)
%
%
% Machine States  = D[P| c v| F] | D[P| v c| F]
% Start State  = [][[]| v c| []]
% Stop State  = [][P| c v| []]

Combinator reconstruction, {{P[c]}}:
%subcode{opsem} include main
% [](c) '= c
% P:c'(c) '= P(c'(;)c)

%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

The small step semantics present for {{langRev}} below work
symmetrically for forward and backward evaluation.

\begin{itemize}
\item 
Basic reduction of an isomorphism. Note that the evaluation leaves the
adjoint of the combinator behind. This will become important when we
reverse the direction of computation.
%% %subcode{opsem} include main
%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>

%subcode{opsem} include main
% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]

\item
Sequencing involves pushing and popping from the Future and Past
continuations:
%subcode{opsem} include main
% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]

\item
Parallel composition, captures the current Future and Past and extends
the parallel context.
%subcode{opsem} include main
% D[P|(left v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
% D[P| (right v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (left v)| F] 
% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (right v)| F]

\item
Similarly for products:
%subcode{opsem} include main
% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]

and symmetrical rules for evaluation along the second branch. 
%subcode{opsem} include main
% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]

The later two rules will be relevant only for reverse execution.

\begin{proposition}[Logical Reversibility]
\label{prop:logrevEE}
{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
\end{proposition}

\begin{proposition}[Groupoid]
\label{prop:groupoidEE}
{{langRevEE}} is a groupoid. 
\end{proposition}

\begin{proposition}
\label{prop:categoryEE}
{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
\end{proposition}

Note: We use `Monoidal Closed' in the sense of Compact Closed. While
the category has an multiplicative tensor that distributes over an
additive tensor, it does not have injections or projections and hence
does not have categorical sums and products.

%%%%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

The operation {{eps}} reverses the direction of a particle by
reversing the world.

Note that we deviate from the categorical definition of {{eta}} and
{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
choice of {{eta}}. This however does not affect us, because we deal
with a symmetric category.

\begin{itemize}
\item Grammar
%subcode{bnf} include main
% Values, v = () | (v, v) | left v | right v | -v
% Isomorphisms, iso &=& ... | eta | eps
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c

\item
Type judgment.
%subcode{opsem} include main
% eta &: 0 <-> (-b) + b :& eps

\item
Operational Semantics.
%subcode{opsem} include main
% D[P| (right v) eps| F]      &|-->&  D{dagger}[F | eps (left (-v))| P]
% D[P| (left (-v)) eps| F]      &|-->&  D{dagger}[F | eps (right v)| P]

Note: there is NO reduction rule for {{eta}}.

\item
The adjoint of a parallel context is defined to be:
%subcode{opsem} include main
% []{dagger} '= []
% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)

\end{comment}

>>>>

We start with the semantics for {{langRev}} defined in Sec
\ref{sec:pi}, which we will systematically rewrite with unification
and then present a ``reverse'' evaluator written in the same style. We
shall then combine these into the semantics for {{langRevEE}}.

\paragraph{Primitive Isomorphsims.}
Previously the reductions of the primitive isomorphisms were specified
in the following form.

{{ iso v_{input pattern} |--> v_{output pattern} }}

\noindent
In the precense of unification, we rewrite them to accept the incoming
substitution and extend it with the appropriate unification.

{{ iso s v' |--> (v_{output pattern}, s[v' <><> v_{input pattern}]) }}

\paragraph{Extending the evaulator with unification.} 
We have rewriten the semantics of Sec. \ref{sec:pi} with
unification. The behaviour of the intepreter has not changed.

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
% <iso, v, cc, s>_{fwd} &|-->& {[iso, v', cc, s']}_{fwd}
% & & where iso s v |--> (v', s')
% <c1(;)c2, v, cc, s>_{fwd} &|-->& <c1, v, Fst cc c2, s>_{fwd}
% {[c1, v, Fst cc c2, s]}_{fwd} &|-->& <c2, v, Snd c1 cc, s>_{fwd}
% {[c2, v, Snd c1 cc, s]}_{fwd} &|-->& {[ c1(;)c2, v, cc,s ]}_{fwd}
% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c1, v, LeftPlus cc c2, s[v' <><> left v]>_{fwd}
% {[ c1, v, LeftPlus cc c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, cc,s ]}_{fwd}
% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c2, v, RightPlus c1 cc, s[v' <><> right v]>_{fwd}
% {[ c2, v, RightPlus c1 cc,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, cc,s ]}_{fwd}
% <c1(*)c2, v', cc, s>_{fwd} &|-->& <c1, v1, LeftTimes cc c2 v2, s'>_{fwd}
% & & where s' = s[v <><> (v1, v2)]
% {[ c1, v1, LeftTimes cc c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightTimes c1 v1 cc, s>_{fwd}
% {[ c2, v2, RightTimes c1 v1 cc,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), cc,s ]}_{fwd}

\paragraph{Backward evaluator.}
Here we have written a ``backward'' evaluator in the same style --
i.e. evaluation {{c v}} in this evaluator is equivalent to evaluating
{{C^{dagger} v}} in the forward evaluator.

%subcode{opsem} include main
%! columnStyle = rclr
%! bck = \triangleleft
% {[iso, v, cc, s]}_{bck} &|-->& <iso, v', cc, s'>_{bck}
% & & where iso^{dagger} s v |--> (v', s')
% <c1, v, Fst cc c2, s>_{bck} &|-->& <c1(;)c2, v, cc, s>_{bck}
% <c2, v, Snd c1 cc, s>_{bck} &|-->& {[c1, v, Fst cc c2, s]}_{bck}
% {[ c1(;)c2, v, cc, s ]}_{bck} &|-->& {[c2, v, Snd c1 cc, s]}_{bck}
% <c1, v, LeftPlus cc c2,s >_{bck} &|-->& <c1(+)c2,left v, cc>_{bck}
% {[c1 (+) c2, v', cc, s ]}_{bck} &|-->& {[ c1, v, LeftPlus cc c2, s[v' <><> left v] ]}_{bck}
% <c2, v, RightPlus c1 cc, s>_{bck} &|-->& <c1(+)c2, right v, cc>_{bck}
% {[c1 (+) c2, v', cc,s ]}_{bck} &|-->& {[ c2, v, RightPlus c1 cc, s[v' <><> right v] ]}_{bck}
% <c1, v1, LeftTimes cc c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), cc, s>_{bck}
% <c2, v2, RightTimes c1 v1 cc, s>_{bck} &|-->& {[ c1, v1, LeftTimes cc c2 v2, s ]}_{bck}
% {[ c1 (*) c2, v, cc, s ]}_{bck} &|-->& {[ c2, v2, RightTimes c1 v1 cc, S' ]}_{bck}
% && where s' = s[v <><> (v1, v2)]

\subsection{Semantics of {{langRevEE}} }

\begin{enumerate}
\item 
To add multiplicative duality, we add the following rules to the set
of primitive isomorphisms:

\begin{enumerate}
\item {{eta* s () |--> ((inv(v), v), s)}}
  where {{v}} is a fresh logic variable.

\item {{eps* s v |--> (1, s[v <><> (inv(v'),v') ])}}
  where {{v'}} is a fresh logic variable.

\end{enumerate}

\item To add negative types we add the following rules to the
  reductions above:

  \begin{enumerate}
  \item

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eps+, v, cc, s>_{fwd} |--> <eps+, left (-v'), cc, s[v <><> right v]>_{bck}
% <eps+, v, cc, s>_{fwd} |--> <eps+, right v', cc, s[v <><> left (-v')]>_{bck}

    \item
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, cc, s>_{bck} |--> <eta+, left (-v'), cc, s[v <><> right v]>_{fwd}
% <eta+, v, cc, s>_{bck} |--> <eta+, right v', cc, s[v <><> left (-v')]>_{fwd}

  \end{enumerate}

\end{enumerate}
